---
title: "Project1_Group5"
author: "Sakshi, Anshita, Shubham, Unnati"
date: "03/01/2022"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## Including all the required libraries

```{r}

library(lubridate)
library(dplyr)
library(dlookr)
library(ggplot2)
library(igraph)
library(reshape)
library(zoo)
library(ggmap)
library(tidyverse)
library(ggraph)
library(tidygraph)
library(ggpubr)
library(sf)
library("wordcloud2")
library(wordcloud)
library(webshot)
library(plotly)
library(forecast)
library(knitr)

#To display map in mac
webshot::install_phantomjs() 

#To display map in Windows
install_phantomjs(version = "2.1.1",
  baseURL = "https://github.com/wch/webshot/releases/download/v0.3.1/",
  force = FALSE)
```

## Loading dataset

```{r}

df_bike_sharing <-  read.csv("hour.csv", header = TRUE, sep = ",")
df_bluebikes_edited <- read.csv("Bluebikes_edited.csv", header = TRUE,
                                strip.white = TRUE, sep = ",")

View(df_bike_sharing)
View(df_bluebikes_edited)

```

## Renaming columns to incorporate sensible column names

```{r}

col_names <- c("ID", "Date", "Season", "Year", "Month", "Hour", "Holiday", "Weekday", "Working_Day",
       "Weather_Situation", "Norm_Temp", "Norm_Feels_Temp", "Norm_Humidity", "Norm_Windspeed",
       "Users_Unregistered", "Users_Registered", "Users_Total" )
colnames(df_bike_sharing) <- col_names

```

## Number of instances and attributes

```{r}
dim(df_bike_sharing)
```

## Checking the datatypes for the columns

```{r}
sapply(df_bike_sharing, class)
```

## Dataset summary

```{r}
summary(df_bike_sharing)
```

## ------------------------- Data Cleaning and Preprocessing ------------------------- ##
## Diagnose dataset to look for missing values

```{r}
diagnose(df_bike_sharing) # no missing values found
```

## Changing attributes, creating new ones for more interpretability

```{r}

df_bike_sharing$Date <- dplyr::case_when(
      substring(df_bike_sharing$Date,1,4) == "2011" ~ 
        sub("2011", "2020",df_bike_sharing$Date),
      substring(df_bike_sharing$Date,1,4) =="2012" ~
        sub("2012", "2021",df_bike_sharing$Date),
      TRUE ~ as.character(df_bike_sharing$Date)
)

# Changing the 'Date' column datatype to Date
df_bike_sharing$Date <- as.Date(df_bike_sharing$Date)

# Extracting the Day from Date and storing in a new column
df_bike_sharing$Day_of_Month <- format(df_bike_sharing$Date, format = "%d")

# Converting Month numerical values to their designated month names
df_bike_sharing$Month <- month.abb[df_bike_sharing$Month]

# Converting Year column from values 0 and 1 to actual year values
# The value 0 represents year 2011 and 1 represents 2012
df_bike_sharing$Year <- ifelse(df_bike_sharing$Year == 0, 2020, 2021)

# Converting Weekday numerical values to their weekday names
df_bike_sharing$Day_of_Week <- dplyr::case_when(
  df_bike_sharing$Weekday == 0 ~ "Sun",
  df_bike_sharing$Weekday == 1 ~ "Mon",
  df_bike_sharing$Weekday == 2 ~ "Tue",
  df_bike_sharing$Weekday == 3 ~ "Wed",
  df_bike_sharing$Weekday == 4 ~ "Thur",
  df_bike_sharing$Weekday == 5 ~ "Fri",
  df_bike_sharing$Weekday == 6 ~ "Sat",
  TRUE ~ as.character(df_bike_sharing$Weekday)
)

# Converting integer values of Hour to HH:MM format
df_bike_sharing$Hour <- sprintf("%02d", df_bike_sharing$Hour)

# Adding Yearly Quarter Column
df_bike_sharing$Quarter = as.yearqtr(df_bike_sharing$Date, format = "%Yq%q")

# Converting numerical Season column to categorical to denote seasons:
# 1 - Winter
# 2 - Spring
# 3 - Summer
# 4 - Fall
df_bike_sharing$Season <- as.character(factor(df_bike_sharing$Season, levels = 1:4,
                                 labels = c("Winter", "Spring", "Summer", "Fall")))

# Converting Weather Situation to its respective assigned values
# 1: Clear, Few clouds, Partly cloudy, Partly cloudy
# 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
# 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
# 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
df_bike_sharing$Weather_Situation <- as.character(factor(df_bike_sharing$Weather_Situation, levels = 1:4,
                                 labels = c("Clear or Partly Cloudy",
                                            "Mist and Cloudy",
                                            "Light Rain or Snow",
                                            "Heavy Rain or Snow")))

```

## Creating new columns to convert normalized values to actual values
## The temperature columns Temp, Feels_like_temp are normalized and calculated using:
## (t - t_min) / (t_max - t_min)
## Temp : t_min = -8, t_max = +39 (Celsius scale)
## Feels_like_temp : t_min = -16, t_max = +50 (Celsius scale)

## Humidity and Windspeed columns are normalized to the scale:
## Humidity - max value of 100
## Windspeed - max value of 67

```{r}

# Converting temperature values to their actual values and storing in new columns
df_bike_sharing$Temp <- (df_bike_sharing$Norm_Temp * (39 + 8)) - 8
df_bike_sharing$Feels_Temp <- (df_bike_sharing$Norm_Temp * (50 + 16)) - 16

# Converting Humidity and Windspeed
df_bike_sharing$Humidity <- df_bike_sharing$Norm_Humidity * 100
df_bike_sharing$Windspeed <- df_bike_sharing$Norm_Windspeed * 67

```

## Fabricating the routes for bike sharing data

```{r}

df_bluebikes_routes_unique <- df_bluebikes_edited %>%
  filter(start.station.name != end.station.name) 

# Reading the start and end locations into a dataframe
df_bluebikes_routes <- dplyr::select(df_bluebikes_routes_unique,
                          c(start.station.name,end.station.name,
                          Source.Longitude,Source.Latitude,
                          Destination.Longitude,Destination.Latitude))

df_bluebikes_routes_dist <- distinct(df_bluebikes_routes)

# Sampling the routes to create random routes for each row of original data
df_random_routes <- sample_n(df_bluebikes_routes_dist, nrow(df_bike_sharing), replace = TRUE)
df_bike_sharing <- cbind(df_bike_sharing, df_random_routes)
ncol(df_bike_sharing)
colnames(df_bike_sharing)[c(25, 26, 27, 28, 29, 30)] <- c("Source", "Destination",
                                                  "Src_Long", "Src_Lat", "Dest_Long", "Dest_Lat")

```

## Viewing the Final Data -- Ready for Visualization

```{r}
View(df_bike_sharing)
```

## ------------------------------- Data Visualization ------------------------------- ##
## Grouped Bar Chart

## This chart helps us to visually compare the monthly bike usage for years 2020 ans 2021

```{r}

df_bike_grouped_bar <- df_bike_sharing %>%
  select(Month, Year, Users_Total) %>%
  group_by(Month, Year) %>%
  summarise(Users = sum(Users_Total))

df_bike_grouped_bar$Month = factor(df_bike_grouped_bar$Month, levels = month.abb)

ggplot(df_bike_grouped_bar, aes(x = Month, y = Users, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  geom_text(aes(label = Users),
            vjust = 0.5, hjust = -0.1, color = "black", size = 1.7, position = position_dodge(0.9)) +
  scale_fill_brewer(palette = "Dark2") +
  ggtitle("Comparison of Monthly Bike Usage for 2020 and 2021") +
  xlab("Months") +
  ylab("Total Number of Users") +
  labs(fill = "Year") +
  scale_x_discrete(limits = ~month.abb) +
  theme(plot.title = element_text(size = 13, face = "bold.italic", color = "black", hjust = 0.5),
        axis.title.x = element_text(size = 7, face = "bold"),
        axis.title.y = element_text(size = 7, face = "bold"),
        axis.text.x = element_text(size = 5, angle = 0),
        axis.text.y = element_text(size = 5, angle = 90),
        legend.position = "right",
        legend.title = element_text(size = 7, face = "bold"),
        legend.text = element_text(size = 7, face = "bold"),
        strip.text = element_text(size = 7))

```

## Box Plot

## This plot shows the weekly distribution of registered and unregistered users

```{r}

df_bike_boxplot <- df_bike_sharing %>%
  select(Weekday, Users_Unregistered, Users_Registered)

df_bike_boxplot_long <- df_bike_boxplot %>% 
  pivot_longer(-Weekday,
               names_to = "Type_of_User",
               values_to = "Users")

ggplot(data = df_bike_boxplot_long, aes(x = as.character(Weekday), y = Users)) +
  geom_boxplot(aes(fill = Type_of_User)) +
  scale_x_discrete(labels = c("0" = "Sunday",
                              "1" = "Monday",
                              "2" = "Tuesday",
                              "3" = "Wednesday",
                              "4" = "Thursday",
                              "5" = "Friday",
                              "6" = "Saturday")) +
  ggtitle("Weekly Distribution of Registered and Unregistered Users") +
  xlab("Weekday") +
  ylab("Number of Users") +
  labs(fill = "Type of Users") +
  theme(plot.title = element_text(size = 13, face = "bold.italic", color = "black", hjust = 0.5),
        axis.title.x = element_text(size = 7, face = "bold"),
        axis.title.y = element_text(size = 7, face = "bold"),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        legend.position = "bottom",
        legend.title = element_text(size = 7, face = "bold"),
        legend.text = element_text(size = 7),
        strip.text = element_text(size = 7)) +
  stat_summary(fun.y = "mean", colour = "darkred", geom = "point",
               shape = 19, size = 3, show.legend = FALSE) 

```

## Heatmap

## This heatmap shows the correlation between Registerd and Unregistered Users with the Weather Attributes

```{r}

# Create a dataframe with numerical columns
df_numeric <- df_bike_sharing %>%
  select(Temp, Feels_Temp, Humidity, Windspeed, Users_Unregistered, Users_Registered)

# Correlation Matrix of all numerical columns
df_bike_heatmap <- round(cor(df_numeric), 2)
df_bike_heatmap <- melt(df_bike_heatmap)

# Plotting heatmap
ggplot(data = df_bike_heatmap, aes(x = X1, y = X2, fill = value)) + 
  geom_tile() +
  geom_text(aes(X1, X2, label = value), color = "black", size = 4) +
  ggtitle("Correlation Heatmap of Users and Weather Attributes") +
  theme_minimal() +
  labs(fill = "Correlation Value") +
  theme(plot.title = element_text(size = 13, face = "bold.italic", hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(size = 7, angle = 90),
        axis.text.y = element_text(size = 7),
        legend.position = "right",
        legend.title = element_text(size = 7, face = "bold"),
        legend.text = element_text(size = 7, face = "bold"))

```

## Bar Plot

## This chart shows the hourly usage of bikes based on different seasons

```{r}

df_bike_bar <- df_bike_sharing %>%
  select(Hour, Season, Users_Total) %>%
  group_by(Hour, Season) %>%
  summarise(Users = ceiling(mean(Users_Total)))

# Plotting bar graph

# Since the x label used in the bar graph is ordinal categorical (hours of the day), hence we have not sorted the bars in decreasing order of their heights.
ggplot(df_bike_bar, aes(x = Hour, y = Users)) +
  geom_bar(stat = "identity", position="dodge", aes(fill = Season)) +
  facet_wrap(~Season, scales = "free_x") +
  geom_text(aes(label = Users), angle = 0, vjust = -0.5,
            color = "black", size = 1.5, angle = 90) +
  scale_fill_brewer(palette = "Dark2") +
  ggtitle("Hourly Usage of Bikes Based on Different Seasons") +
  xlab("Hour of the Day") +
  ylab("Number of Users") +
  theme(plot.title = element_text(size = 13, face = "bold.italic", color = "black", hjust = 0.5),
        axis.title.x = element_text(size = 7, face = "bold"),
        axis.title.y = element_text(size = 7, face = "bold"),
        axis.text.x = element_text(size = 5, angle = 90),
        axis.text.y = element_text(size = 5),
        legend.position = "right",
        legend.title = element_text(size = 7, face = "bold"),
        legend.text = element_text(size = 7, face = "bold"),
        strip.text = element_text(size = 7))

```

## Line Chart

## This chart shows the quarterlu usage of bikes based on weather situation
## We are also predicting the number of users for next quarters

```{r Line Chart}

df_bike_line <- df_bike_sharing %>%
  group_by(Quarter, Weather_Situation) %>%
  summarise(Users = ceiling(mean(Users_Registered)))

ggplot(df_bike_line, aes(x = Quarter,  y = Users)) +
  geom_line() +
  geom_smooth(aes(color = Weather_Situation,
                  fill = Weather_Situation), method = "lm") +
  geom_point() +
  ylim(0, 1000) +
  facet_wrap(~ Weather_Situation, scales = "free_x") +
  geom_forecast(stat = "forecast", position = "identity",
                colour = "lightblue", showgap = FALSE) +
  geom_text(aes(label = Users),
            vjust = -2.5, color = "black", size = 1.7) +
  scale_fill_brewer(palette = "Dark2") +
  ggtitle("Quarterly Usage of Bikes Based on Weather Situation with Future Prediction") +
  xlab("Year and Quarters") +
  ylab("Number of Users") +
  scale_x_yearqtr(format = "%YQ%q") +
  theme(plot.title = element_text(size = 13, face = "bold.italic", color = "red", hjust = 0.5),
        axis.title.x = element_text(size = 7, face = "bold"),
        axis.title.y = element_text(size = 7, face = "bold"),
        axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(colour = "blue", size = 7, face = "bold"),
        strip.text = element_text(size = 7))

```

## Network Graph

## The graph is showcasing the 20 minimally used routes.

```{r}

# Selecting the source and destination and summarizing the total users
plot_net <- dplyr::select(df_bike_sharing, c(Source, Destination, Users_Total))

plot_net_filtered <- plot_net %>%
  group_by(Source, Destination) %>%
  summarise(Users_Avg = mean(Users_Total)) %>%
  arrange(Users_Avg) %>%
  head(20)

# Creating a network
grph_net <- graph.data.frame(plot_net_filtered[c(1, 2, 3)], directed = TRUE)

V(grph_net)$size <- plot_net_filtered$Users_Avg

# Plotting the network
plot(grph_net,
     layout = layout.auto,
     vertex.size = 10,
     vertex.color = "blue",
     vertex.label.cex = 0.8,
     vertex.label.color = "black",
     edge.arrow.size = 0.1,
)
title(main =list("Rarely Used Routes", cex=1.5))
```

## Map View of stations

##The map shows network of various bike stations and relative locations in Boston

```{r}

# Selecting the source and destination with respective longitude and latitude columns
plot_map_filtered <- df_bike_sharing %>%
  dplyr::select(c(Source, Destination, Src_Long,
                  Src_Lat, Dest_Long, Dest_Lat, Users_Total)) %>%
  group_by(Source, Destination, Src_Long,
                  Src_Lat, Dest_Long, Dest_Lat) %>%
  summarise(Users = round(mean(Users_Total), 0))

# Using Google API to plot the mapview
api_key <- register_google(key = "AIzaSyCO_xreF1k7Gx-grRe5Dzz17BdzRca5398")
map_canvas <- get_map(c(left = min(plot_map_filtered$Src_Long), 
                  bottom = min(plot_map_filtered$Src_Lat), 
                  right = max(plot_map_filtered$Dest_Long), 
                  top = max(plot_map_filtered$Dest_Lat)),
                  maptype = "satellite",
                  source = "google",
                  zoom = 10)

# Plotting using ggmap
ggmap(map_canvas, darken = c(0.6, "white")) + 
  geom_segment(data = plot_map_filtered,
               aes(x = Src_Long, 
                   y = Src_Lat,
                   xend = Dest_Long,
                   yend = Dest_Lat,
                   alpha = sqrt(Users)),
               color = "#000000") +
  coord_cartesian() +
  scale_alpha(range = c(0.0001, .5)) +
  geom_point(data = plot_map_filtered %>% 
               group_by(longitude = Src_Long,
                        latitude = Src_Lat) %>%
               summarize(rides = sum(Users)),
             aes(x = longitude, y = latitude, size = rides),
             color = "#009900", alpha = .4) + 
  scale_size_continuous(range(4, 100)) +
  scale_color_viridis_c() + 
  scale_fill_viridis_c() + 
  theme_nothing()

# Saving the html output as png in local drive
ggsave(filename = "station-network.jpg", width = 8, units = "in")

```

## Word Cloud

## This visualization is displaying the routes, the size of the text depends on the number of users
   travelling on the route. 

```{r}

# Aggregating Total Number of users on a particular route using Group by

plot_cloud_filtered <- df_bike_sharing %>%
  dplyr::select(c(Source, Destination, Users_Total)) %>%
  group_by(Source, Destination) %>%
  summarise(Users = ceiling(mean(Users_Total)))

# Concatenating source and destination
plot_cloud_filtered$concat <- paste(plot_cloud_filtered$Source,
                                    plot_cloud_filtered$Destination,
                                    sep = " --> ")

set.seed(1234)
word_cloud <- wordcloud2(data = plot_cloud_filtered[c(4,3)],
                        size = 0.5, color = 'random-dark', gridSize = 15,
                        shape = "diamond", ellipticity = 0.2,
                        fontWeight = "normal", shuffle = FALSE) 
word_cloud

```